Experiences with planning for natural language generation

### 2008-06-20 trunk/Papers/experiences/domains.tex
small changes to the planning in NLG section
pictures
details on the give problem

In this section, we present the two domains: first the sentence
generation domain, then the instruction giving domain.

@@

\begin{figure}
  \centering
  xxxx
  \caption{The example grammar.}
  \label{fig:white-rabbit-sleeps-grammar}
\end{figure}

\begin{figure}
  \centering
  xxxx
  \caption{Derivation of ``The white rabbit sleeps.''}
  \label{fig:white-rabbit-sleeps-deriv}
\end{figure}

@@

The object of the GIVE Challenge (``Generating Instructions in Virtual
Environments''; Koller et al.\
\citeyear{alexander07:_shared_task_propos}) is to build an NLG system
which is able to give natural-language instructions which will guide a
human user in performing some task in a virtual environment.  A map of
an example world is shown in Fig.~\ref{fig:give-development-world};
instructions for the first steps could be ``turn left; press the
button; turn right and walk through the door;'' and so forth.  From an
NLG perspective, GIVE makes for an interesting challenge because it is
a theory-neutral task that exercises all components of an NLG system,
and emphasizes the study of communication in a (simulated) physical
environment.  It also has the advantage that the user and the NLG
system can be physically in different places, as long as the 3D client
and the NLG system are connected over a network.  This makes it
possible to evaluate GIVE NLG systems on a large scale over the
Internet.  The first GIVE evaluation will take place in late 2008;
currently seven research teams from five countries are working on
developing systems to participate in the challenge.

@@

\todo{explain world? -- yes, definitely}

One crucial component of a GIVE system is a planner which will compute
a domain plan that consists of the actions the user has to execute to
achieve the goal.  For these purposes, the virtual world is
discretized into a planning problem in which each object has a
symbolic position and orientation.  The first few action instances in
the example domain could be $turn-left(north,west)$, $move(pos-3-2,
pos-2-2, west)$, $manipulate-b_1-off-on(pos-2-2)$,
$turn-right(west,north)$, \todo{complete this} and so on.  Under the
current implementation of the planning problem, the shortest plan for
getting the player from the initial state to the goal (in which they
are holding the trophy) has 104 steps, most of which are ``move''
actions.  The bulk of the GIVE problem is therefore a variant of the
Gridworld problem, which also involves finding a route through a world
with discrete positions -- but with the need to press buttons in the
right order and with somewhat more complicated room shapes.

It is interesting to note that although the task of a GIVE NLG system
could be seen as mapping a domain plan to a sequence of NL
instructions, this mapping is not trivial.  It would be
\emph{possible} to express the last part of the above example plan as
``turn right; walk forward; walk forward; turn right; walk forward;
walk forward; turn left; walk forward'', but this is a very clumsy way
of phrasing the instruction ``turn around and walk through the door''
that will frustrate users and lead to bad evaluation results.  In
other words, it is important to aggregate multiple movement actions
into a single NL instruction.  On the other hand, it may be necessary
to generate instructions that don't correspond to domain actions.  For
instance, it will be necessary to find a correct referring expression
for $b_1$ in order to express the action
$manipulate-b_1-off-on(pos-2-2)$.  The interaction between referring
and instructing can go much deeper than that.  Imagine, for example,
that the user just entered the top right room, and we want to instruct
them to press $b_4$.  At that point in time, the user can see six
different buttons, and generating a single instruction of the form
``push X'' is difficult.  However, an instruction sequence like ``walk
forward; turn left; now press the button right in front of you'' will
be successful and natural.  Here the first two instructions do double
duty by preparing the simple referring expression ``the button right
in front of you'', which is only valid after the user's position and
orientation changed.  A generation system that is supposed to generate
such instruction sequences must be tightly integrated with the planner
in order to exploit the simultaneous effect of the first two
instructions on the world and on the communicative situation.

>>>>

In this section, we present our planning domains: the sentence
generation domain and the instruction giving domain.

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{pic-grammar}
  \caption{The example grammar.}
  \label{fig:white-rabbit-sleeps-grammar}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{pic-derivation}
  \caption{Derivation of ``The white rabbit sleeps.''}
  \label{fig:white-rabbit-sleeps-deriv}
\end{figure}

@@

The object of the GIVE Challenge (``Generating Instructions in Virtual
Environments''; Koller et al.\
\citeyear{alexander07:_shared_task_propos}) is to build an NLG system
which is able to give natural-language instructions which will guide a
human user in performing some task in a virtual environment.  From an
NLG perspective, GIVE makes for an interesting challenge because it is
a theory-neutral task that exercises all components of an NLG system,
and emphasizes the study of communication in a (simulated) physical
environment.  It also has the advantage that the user and the NLG
system can be physically in different places, as long as the 3D client
and the NLG system are connected over a network.  This makes it
possible to evaluate GIVE NLG systems on a large scale over the
Internet.  The first GIVE evaluation will take place in late 2008;
currently eight research teams from five countries are working on
developing systems to participate in the challenge.

@@

A map of an example world is shown in
Fig.~\ref{fig:give-development-world}.  In this world, the user's task
is to pick up a trophy in the top left room.  The trophy is hidden in
a safe behind a picture, and the user must therefore first move the
picture away and open the safe by pushing a sequence of buttons (the
small square boxes on the walls) in the correct order.  In order to
get to all these buttons, they must also open the lower door and
deactivate the alarm tile by pushing further buttons.  To simplify
both the planning and the NLG task, the world is discretized into
tiles of equal size; the user can turn by 90 degree steps in either
direction, and can move from the center of one tile to the center of
the next.  Some actions that the user can take are shown in
Fig.~\ref{fig:give-planning}.  The first few steps of one shortest
plan to achieve the goal are:
\begin{enumerate}
\item $\mathsf{turn-left}(\mathsf{north}, \mathsf{west})$
\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$
\item $\mathsf{manipulate-b1-off-on}(\mathsf{pos\_5\_2})$
\item $\mathsf{turn-right}(\mathsf{west}, \mathsf{north})$
\end{enumerate}

\begin{figure}
  \centering
  xxxxx
  \caption{Excerpt from the GIVE planning problem.}
  \label{fig:give-planning}
\end{figure}

This shortest plan consists of 108 action instances, most of which are
``move'' actions.  The bulk of the GIVE problem is therefore a variant
of the Gridworld problem, which also involves finding a route through
a world with discrete positions -- but with the need to press buttons
in the right order and with somewhat more complicated room shapes.  In
this example, the NLG system could generate an instruction sequence
starting with ``turn left; walk forward; press the button; turn right;
walk forward; walk forward; turn right; walk forward; walk forward;
turn left; walk forward''.  However, this would be clumsy and boring,
and it is to be expected that frustrated users will quickly cancel an
interaction with such a system, resulting in a bad evaluation score.
Having just pressed the button, it would be much better to say ``turn
around and walk through the door''.  This means that the mapping from
action instances to natural language instructions is not trivial.
Conversely, it may also be necessary to express a single planning step
with several instructions.  Having just entered the top left room, it
may be easier for the user to understand the instruction ``walk to the
center of the room; turn right; now press the green button in front of
you'' than the instruction ``press the green button on the wall to
your right''.  To plan the referring expression ``the green button in
front of you'' at a time when the button is not yet in front of the
user, the NLG system must keep track of the hypothetical changes in
what objects are visible to the user.  Thus the acts of referring and
instructing, and the modules for planning and NLG, must be integrated
tightly.
